---
title: "A Stasticical Analysis of Equine Corneal Anesthesia"
author: 
   - Zach Goldstein, CSU College of Natural Sciences
   - Dr. Ann Hess, Statistics
   - Dr. Rachel Hector, CSU Vet Hospital
date: "June 27, 2025"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
# load libraries
library(knitr)
library(tidyverse)
library(openxlsx)
library(ggplot2)
library(janitor)
library(dplyr)
library(scales)
library(lme4)
library(lmerTest) 
library(emmeans)
library(kableExtra)
library(broom.mixed)
library(broom)
```

```{r data-import, include=FALSE}
# Summarized data
df_summ <- read.xlsx(
  "Ropivacaine_Dexmed_Data_AH.xlsx", sheet = "Summarized"
) %>%
  rename(
    horse                     = Horse,
    treatment                 = Treatment,
    period                    = Period,
    minimum_ctt_cm            = `Minimum.CTT.(cm)`,
    duration_of_min_ctt_min   = `Duration.of.Min.CTT.(min)`,
    maximum_corneal_pressure  = Maximum.corneal.pressure,
    total_anesthesia_time_min = `Total.Anesthesia.Time.(min)`
  ) %>%
  mutate(
    horse     = factor(horse),
    treatment = factor(treatment, levels = c("DR", "R", "S")),
    period    = factor(period)
  )

# Add horse_period
 df_summ2 <- df_summ %>%
  mutate(horse_period = interaction(horse, period, sep = "_"))

# Hemorrhage score data
df_hem_raw <- read.xlsx(
  "Ropivacaine_Dexmed_Data_AH.xlsx", sheet = "HemScore"
) %>%
  clean_names()

df_hem <- df_hem_raw %>%
  mutate(
    horse     = factor(horse, levels = levels(df_summ2$horse)),
    treatment = factor(treatment, levels = levels(df_summ2$treatment)),
    hem_score = replace_na(hemorrhage_score, 0)
  ) %>%
  select(horse, treatment, hem_score)

df_summ4 <- df_summ2 %>%
  left_join(df_hem, by = c("horse", "treatment")) %>%
  mutate(hem_score = replace_na(hem_score, 0))

# Full time-course data
df_full <- read.xlsx(
  "Ropivacaine_Dexmed_Data_AH.xlsx", sheet = "Full"
)

df_full_long <- df_full %>%
  pivot_longer(
    cols = -c(Horse, Treatment, Variable),
    names_to  = "time_label",
    values_to = "value"
  ) %>%
  mutate(time_min = as.numeric(time_label))
```

# Background and Motivation

- Corneal procedures in horses, like fixing corneal flaps or removing foreign bodies, are often done with the horse standing rather than under full anesthesia. This is because **general anesthesia carries about a 1% risk of serious complications during recovery.**  
- To keep the horse comfortable and safe, we need a reliable way to **numb the surface of the eye locally,** so they don’t feel pain but can still stand on their own.  
- We routinely inject **ropivacaine** under the tissue around the eye to block sensation for several hours, but many clinicians add **dexmedetomidine** (a mild sedative) in the hope that it will extend that numbness even longer.  
- It isn’t yet proven whether adding dexmedetomidine really makes the cornea stay numb longer -- or whether it changes other safety factors, like causing more bleeding where we inject. 

# Background and Motivation

- To find out, a controlled crossover study was ran, measuring four things:  
  1. **How long the eye stayed numb** (Total Anesthesia Time, or TTA)  
  2. **How deep the numbness went** (minimum corneal touch threshold)
  3. **How long the deepest numbness lasted** (duration of minimum CTT)  
  4. **Any bleeding at the injection site** (hemorrhage score)  


#  **What is Corneal Touch Threshold (CTT)?**  
> CTT was measured with a Cochet–Bonnet esthesiometer: a very fine, retractable nylon filament that is gently touched to the center of the cornea. The longer the filament, the less force it applies; the shorter it is, the more pressure it puts on the eye. If the horse blinks, the filament length is recorded in centimeters as the CTT. A CTT of 6 cm (maximum length) means the cornea is fully sensitive; a CTT of 0 cm (shortest length) means the horse felt nothing at any pressure: the cornea is maximally numb.

# Study Design

- Ten healthy adult horses were enrolled (we excluded one eye due to cataract) in a **two-period crossover**.  Each horse’s two eyes took turns being the treatment eye vs. saline control, with assignments randomized and a one-week break in between.  
- Before each session the horse was lightly sedated with xylazine and recorded baseline measurements:  
  - **CTT** using the filament method above  
  - **Corneal pressure** using a small tonometer  
- After injecting the drug under the tissue around the cornea, these measurements were reported **every 10 minutes** for up to 2 hours (and then every 30 minutes if numbness persisted), until the CTT filament returned to its original length on two checks in a row.   
- **Outcomes**: total anesthesia time (TTA), minimum CTT, duration of minimum CTT, maximum pressure; hemorrhage score (0–4); ambient humidity recorded.

# Exploratory Data

```{r eda-spaghetti, echo=FALSE}
# TTA spaghetti plot by horse and treatment
ggplot(df_summ2, aes(
  x         = treatment,
  y         = total_anesthesia_time_min,
  group     = horse,
  color     = horse
)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Total Anesthesia Time by Treatment (Each Horse)",
    x     = "Treatment",
    y     = "TTA (min)",
    color = "Horse"
  ) +
  theme_minimal(base_size = 12)
```

# Analyzing Total Anesthesia Time

- **Why a mixed‐effects model?**  
  - **Repeated measures**: each horse contributes two eyes × two periods -> non-independence  
  - **Horse as random effect**: captures between‐horse variability in baseline and response  

- **Model specification**  
  - **Fixed effect**: Treatment (DR, R, S)  
  - **Random intercept**: Horse  

- **Variance stabilization**  
  - **Square‐root transform** on TTA to address right skew and heteroscedasticity  

- **Inference**  
  - **Type III F‐test** (Satterthwaite’s approximation) for overall treatment effect  
  - **emmeans** on the Square-Root scale, back‐transformed to original TTA (min) for interpretability  
  - **Paired Wilcoxon signed‐rank test** (DR vs R) as nonparametric confirmation  

# Analyzing Total Anesthesia Time
## Summary Statistics

```{r tta-summary-pres, echo=FALSE}
df_summ2 %>%
  group_by(treatment) %>%
  summarise(
    N      = n(),
    Mean   = mean(total_anesthesia_time_min),
    SD     = sd(total_anesthesia_time_min),
    Min    = min(total_anesthesia_time_min),
    Median = median(total_anesthesia_time_min),
    Max    = max(total_anesthesia_time_min)
  ) %>%
  kable(digits = 1,
        caption = "Summary Statistice for Total Anesthesia Time (Units: $mins$) Response"
        ) %>%
  kable_styling(full_width = FALSE)
```

# Analyzing Total Anesthesia Time

## Mixed Model (Square-Root-Transformed TTA)

```{r tta-sqrt-model-pres, echo=FALSE}
m_tta_sqrt_hp <- lmer(
  sqrt(total_anesthesia_time_min) ~ treatment +
    (1 | horse),
  data = df_summ4,
  REML = TRUE
)

# Type III ANOVA
anova_tbl_sqrt_hp <- as.data.frame(anova(m_tta_sqrt_hp))
anova_tbl_sqrt_hp %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>%  
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for sqrt(Total Anesthesia Time)",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

# Analyzing Total Anesthesia Time

## Estimated Marginal Means

```{r tta-emmeans-pres, echo=FALSE}
#running emmeans
emm_sqrt_tta_hp <- emmeans(m_tta_sqrt_hp, pairwise ~ treatment, adjust = "tukey")

emm_sqrt_tta_hp$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Sqrt Transformed Total Anesthesia Time (Units: $min$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

# Analyting Total Anesthesia Time

## Estimated Marginal Means
```{r emmip-plot, fig.width = 6, fig.height = 3}
# emmip plot
emmip(emm_sqrt_tta_hp, ~ treatment, CI = TRUE, type = "response") +
aes(color = treatment, fill = treatment) +
  scale_color_manual(                          
    values = c(
      DR = "steelblue",   
      R  = "orange",  
      S  = "#009E73"  
    )
  ) +
  scale_fill_manual(                          
    values = c(
      DR = "steelblue",   
      R  = "orange",  
      S  = "#009E73"  
    )
  )
```





# Conclusions

- Overall effect: sqrt(TTA) mixed model highly significant (Type III F(2, 29) = 31.09, p = 5.9e-8).  
- DR vs R: DR increased sqrt(TTA) by 3.42 minutes (SE = 1.26; t = 2.72; p = 0.028). DR vs S: +8.43 minutes (SE = 1.11; t = 7.58; p = 6.9e-8).  
- R vs S: R exceeded S by 5.01 minutes (SE = 1.08; t = 4.63; p = 0.0002).
- Model diagnostics show acceptable fit after square-root‐transform and random effects.

# Analyzing Total Anesthesia Time

## Wilcoxon Test (DR vs R)

```{r tta-wilcox-pres, echo=FALSE}
#paired DR vs R on df_summ2, ensuring complete pairs
cts_tta <- df_summ2 %>%
  filter(treatment %in% c("DR", "R")) %>%
  select(horse, treatment, total_anesthesia_time_min) %>%
  pivot_wider(
    names_from = treatment,
    values_from = total_anesthesia_time_min
  ) %>%
  filter(!is.na(DR) & !is.na(R))

#run Wilcoxon signed‐rank test with Hodges–Lehmann CI
wil_tta <- wilcox.test(
  cts_tta$DR,
  cts_tta$R,
  paired   = TRUE,
  exact    = FALSE,
  conf.int = TRUE,
  conf.level = 0.95
)

#summarize test results
tibble(
  Statistic    = wil_tta$statistic,
  `HL Estimate`= wil_tta$estimate,
  `95% Lower`  = wil_tta$conf.int[1],
  `95% Upper`  = wil_tta$conf.int[2],
  `p-value`    = wil_tta$p.value
) %>%
  kable(
    caption = "Wilcoxon Test: Total Anesthesia Time (DR vs R)",
    digits  = c(0,0,0,0,4)
  ) %>%
  kable_styling(full_width = FALSE)
``` 

## Conclusions 

- Wilcoxon signed‐rank (DR vs R), however, does not supports DR > R (p = 0.076).
- *Note: The paired Wilcoxon signed-rank test for DR vs R (Statistic = 38; HL Estimate = 105; 95% CI = -102 to 225; p = 0.0756) was not significant. Because it only considers paired differences and ignores between-horse variability, its power is lower than the mixed model’s, which uses all observations and models horse as a random effect.*  

# Full Time Course For A Single Horse

```{r full-timecourse-pres, echo=FALSE, warning = FALSE, fig.width = 6, fig.height = 3, fig.show = 'hold', out.height = '0.5\\textheight',out.width = '0.8\\textwidth'}

#selecting horse and treatment
sel_horse <- "4"
sel_trt   <- "DR"

#making subset for one horse (and changing CTT to long form)
df_one_ctt <- df_full %>%
  filter(Horse == sel_horse, Treatment == sel_trt, Variable == "CTT (cm)") %>%
  pivot_longer(
    cols = c('Time.-.Baseline', matches("^[0-9]+$")),
    names_to  = "time_label",
    values_to = "ctt_cm"
  ) %>%
  mutate(
    #parse whatever digits are in the label
    time_min = parse_number(time_label),
    #replace that NA with 0
    time_min = replace_na(time_min, 0)
  )

#plotting
ggplot(df_one_ctt, aes(x = time_min, y = ctt_cm)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = paste0("CTT (cm) Over Time: Horse ", sel_horse, " (", sel_trt, ")"),
    x     = "Time (min)",
    y     = "CTT (cm)"
  ) +
  theme_minimal(base_size = 14)

#associated table
#pull the matching summarized row into a new object
example_summary_matched2 <- df_summ2 %>%
  filter(horse == sel_horse, treatment == sel_trt) %>%
  select(
    minimum_ctt_cm,
    duration_of_min_ctt_min,
    total_anesthesia_time_min
  )

#pivot longer so we have Metric / Value
example_summary_matched2_long <- example_summary_matched2 %>%
  pivot_longer(
    cols      = everything(),
    names_to  = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = recode(Metric,
      minimum_ctt_cm            = "Minimum CTT (cm)",
      duration_of_min_ctt_min   = "Duration of Min CTT (min)",
      total_anesthesia_time_min = "Total Anesthesia Time (min)"
    )
  )

#print it
example_summary_matched2_long %>%
  kable(
    col.names = c("Metric", "Value"),
    caption   = paste0(
      "Summarized Metrics for Horse ",
      sel_horse, " (Treatment ", sel_trt, ")"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE,
    font_size         = 8,
    latex_options = c("hold_position", "scale_down")
  )
```

# Full Time Course For A Single Horse (ALL Treatments)
```{r}

# grab CTT for this one horse, all treatments
df_one_horse_ctt <- df_full_long %>%
  filter(
    Horse    == sel_horse,
    Variable == "CTT (cm)"
  ) %>%
  mutate(
    time_min = parse_number(time_label),
    time_min = replace_na(time_min, 0)
  )

ggplot(df_one_horse_ctt, aes(x = time_min, y = value, color = Treatment)) +
  geom_line(size = 1, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(               
    values = c(
      DR = "steelblue",   # a rich orange
      R  = "orange",   # a strong blue
      S  = "#009E73"    # a deep green
    )
  ) +
  labs(
    title = paste0("CTT Over Time: Horse ", sel_horse, " Across Treatments"),
    x     = "Time (min)",
    y     = "CTT (cm)",
    color = "Treatment"
  ) +
  theme_minimal(base_size = 14)
```

# Exploratory Hemorrhage Data

```{r eda-hemorrhage, echo=FALSE, fig.width = 6, fig.height = 3, fig.show = 'hold', out.height = '0.5\\textheight',out.width = '0.8\\textwidth'}
# Hemorrhage score distribution by treatment
ggplot(df_summ4, aes(x = treatment, fill = factor(hem_score))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer("Hem Score", palette = "Reds") +
  labs(
    x = "Treatment",
    y = "Proportion of Eyes",
    title = "Distribution of Hemorrhage Scores by Treatment"
  ) +
  theme_minimal()
```

## Wilcoxon Signed-Rank DR vs R test
```{r wilxoc-hem, fig.width = 6, fig.height = 3, fig.show = 'hold', out.height = '0.5\\textheight',out.width = '0.8\\textwidth'}
#pivot to wide for DR and R
wide_hem <- df_summ4 %>%
  filter(treatment %in% c("DR", "R")) %>%
  select(horse, treatment, hem_score) %>%
  pivot_wider(names_from = treatment, values_from = hem_score)

#run paired Wilcoxon test
wil_hem <- wilcox.test(wide_hem$DR, wide_hem$R, 
                       paired = TRUE, exact = FALSE,
                       conf.int = TRUE,
                       conf.level = 0.95)


#summarize results
tibble(
  Statistic      = wil_hem$statistic,
  `H‐L Estimate` = wil_hem$estimate,
  `95% Lower`    = wil_hem$conf.int[1],
  `95% Upper`    = wil_hem$conf.int[2],
  `p‐value`      = wil_hem$p.value
) %>%
  kable(
    caption = "Wilcoxon Signed-Rank Test with Hodges–Lehmann CI",
    digits  = c(0, 0, 0, 0, 4)
  ) %>%
  kable_styling(full_width = FALSE)
```
