---
title: "Model Tests"
author: "Zach Goldstein"
date: "2025-05-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = FALSE, warning = FALSE)
set.seed(12345)
```

```{r package-setup, include = FALSE}
#Installing Packages
library(tidyverse)
library(openxlsx)
library(ggplot2)
library(janitor)
library(dplyr)
library(scales)
library(lme4)
library(lmerTest) 
library(emmeans)
library(kableExtra)
library(broom.mixed)
library(broom)


```

```{r}
#loading in package

df_summ <- read.xlsx("Ropivacaine_Dexmed_Data_AH.xlsx", sheet = "Summarized") %>%
  rename(
    horse                     = Horse,
    treatment                 = Treatment,
    period                    = Period,
    minimum_ctt_cm            = `Minimum.CTT.(cm)`,
    duration_of_min_ctt_min   = `Duration.of.Min.CTT.(min)`,
    maximum_corneal_pressure  = Maximum.corneal.pressure,
    total_anesthesia_time_min = `Total.Anesthesia.Time.(min)`
  ) %>%
  mutate(
    horse     = factor(horse),
    treatment = factor(treatment, levels = c("DR", "R", "S")),
    period    = factor(period)
  )
```

```{r, results = 'hide'}

df_summ3 <- df_summ %>%
  mutate(
    horse_period = interaction(horse, period, sep = "_")
  )
```

# 7. Adding in HemScore
```{r}
#Load the hemorrhage‐score sheet, keeping only horse, treatment, and score
df_hem <- read.xlsx("Ropivacaine_Dexmed_Data_AH.xlsx", sheet = "HemScore") %>%
  clean_names() %>%
  #Turn the numeric horse ID into the same factor as in df_summ3:
  mutate(
    horse     = factor(horse, levels = levels(df_summ3$horse)),
    treatment = factor(treatment, levels = levels(df_summ3$treatment)),
    hem_score = replace_na(hemorrhage_score, 0)
  ) %>%
  select(horse, treatment, hem_score)
#Left‐join onto df_summ3 by just horse & treatment
df_summ4 <- df_summ3 %>%
  left_join(df_hem, by = c("horse", "treatment")) %>%
  #now period is untouched from df_summ3, and horse_period can be generated:
  mutate(
    period        = factor(period),
    horse_period  = interaction(horse, period, sep = "_")
  )
```

# 1. Response = total_anesthesia_time_min

## 1a. Summary Table

```{r}
######
# Response = Total Anesthesia Time
######

#summary table

sum_tta <- df_summ %>%
  group_by(Treatment = treatment) %>%
  summarize(
    N = n(),
    Mean = mean(total_anesthesia_time_min, na.rm = TRUE),
    SD = sd(total_anesthesia_time_min, na.rm = TRUE),
    Min = min(total_anesthesia_time_min, na.rm = TRUE),
    Median = median(total_anesthesia_time_min, na.rm = TRUE),
    Max = max(total_anesthesia_time_min, na.rm = TRUE),
    .groups = "drop"
  )

sum_tta %>%
  kable(
    digits = 2,
    format = "latex",
    caption = "Summary of Total Anesthesia Time (Units: $min$) by Treatment",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "scale_down"))
```

## 1b. Fit LMM

```{r, results = 'hide'}
#fitting model

m_tta <- lmer(
  total_anesthesia_time_min ~ treatment + period + (1 | horse),
  data = df_summ
)
summary(m_tta)
```

## 1c. Global F Test

```{r}

# capture the ANOVA table as a data.frame
anova_m_tta <- as.data.frame(anova(m_tta))

anova_m_tta %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>% 
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for Duration of Total Anesthesia Time",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 1d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_tta <- emmeans(m_tta, pairwise ~ treatment, adjust = "tukey")

emm_tta$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Total Anesthesia Time (Units: $min$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 1e. Diagnostic Plots

```{r}
#residual v fitted plot
resid_tta <- tibble(
  fitted = fitted(m_tta),
  resid  = resid(m_tta)
)

ggplot(resid_tta, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Total Anesthesia Time",
    subtitle = expression(paste("Response Units: mins"))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_tta)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Total Anesthesia Time",
    subtitle = expression(paste("Response Units: mins")),
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
```

# 2. Response = minimum_ctt_cm

## 2a. Summary Table

```{r}
######
# Response = Minimum CTT (cm)
######

#summary table

sum_mcct <- df_summ %>%
  group_by(Treatment = treatment) %>%
  summarize(
    N = n(),
    Mean = mean(minimum_ctt_cm, na.rm = TRUE),
    SD = sd(minimum_ctt_cm, na.rm = TRUE),
    Min = min(minimum_ctt_cm, na.rm = TRUE),
    Median = median(minimum_ctt_cm, na.rm = TRUE),
    Max = max(minimum_ctt_cm, na.rm = TRUE),
    .groups = "drop"
  )

sum_mcct %>%
  kable(
    digits = 2,
    format = "latex",
    caption = "Summary of Minimum CTT (Units: $cm$) by Treatment",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "scale_down"))
```

## 2b. Fit LMM

```{r, , results = 'hide'}
#fitting model

m_mcct <- lmer(
  minimum_ctt_cm ~ treatment + period + (1 | horse),
  data = df_summ4
)
summary(m_mcct)
```

## 2c. Global F Test

```{r}

# capture the ANOVA table as a data.frame
anova_tbl_mcct <- as.data.frame(anova(m_mcct))

anova_tbl_mcct %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>% 
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for Minimun CCT",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 2d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_mcct <- emmeans(m_mcct, pairwise ~ treatment, adjust = "tukey")

emm_mcct$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Minimum CCT (Units: $cm$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 2e. Wilcoxon Signed Rank Test
```{r}
# pivot to wide for DR and R
wide_minctt <- df_summ %>%
  filter(treatment %in% c("DR", "R")) %>%
  select(horse, treatment, minimum_ctt_cm) %>%
  pivot_wider(names_from = treatment, values_from = minimum_ctt_cm)

# Run paired Wilcoxon test
wil_mcct <- wilcox.test(wide_minctt$DR, wide_minctt$R, 
                       paired = TRUE, exact = FALSE,
                       conf.int = TRUE,
                       conf.level = 0.95)


#summarize results
tibble(
  Statistic      = wil_mcct$statistic,
  `H‐L Estimate` = wil_mcct$estimate,
  `95% Lower`    = wil_mcct$conf.int[1],
  `95% Upper`    = wil_mcct$conf.int[2],
  `p‐value`      = wil_mcct$p.value
) %>%
  kable(
    caption = "Wilcoxon Signed-Rank Test with Hodges–Lehmann CI",
    digits  = c(0, 0, 0, 0, 4)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 2f. Diagnostic Plots

```{r}
#residual v fitted plot
resid_mcct <- tibble(
  fitted = fitted(m_mcct),
  resid  = resid(m_mcct)
)

ggplot(resid_mcct, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Minimum CTT",
    subtitle = expression(paste("Response Units: cm"))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_mcct)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Minimum CTT",
    subtitle = expression(paste("Response Units: cm")),
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
```

# 3. Response = duration_of_min_ctt_min

## 3a. Summary Table

```{r}
######
# Response = Minimum CTT (cm)
######

#summary table

sum_dmcct <- df_summ %>%
  group_by(Treatment = treatment) %>%
  summarize(
    N = n(),
    Mean = mean(duration_of_min_ctt_min, na.rm = TRUE),
    SD = sd(duration_of_min_ctt_min, na.rm = TRUE),
    Min = min(duration_of_min_ctt_min, na.rm = TRUE),
    Median = median(duration_of_min_ctt_min, na.rm = TRUE),
    Max = max(duration_of_min_ctt_min, na.rm = TRUE),
    .groups = "drop"
  )

sum_dmcct %>%
  kable(
    digits = 2,
    format = "latex",
    caption = "Summary of Duration of Minimum CTT (Units: $min$) by Treatment",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "scale_down"))
```

## 3b. Fit LMM

```{r, results = 'hide'}
#fitting model

m_dmcct <- lmer(
  duration_of_min_ctt_min ~ treatment + period + (1 | horse),
  data = df_summ
)
summary(m_dmcct)
```

## 3c. Global F Test

```{r}
# capture the ANOVA table as a data.frame
anova_tbl_dmcct <- as.data.frame(anova(m_dmcct))

anova_tbl_dmcct %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>% 
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for Duration of Minimun CCT",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 3d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_dmcct <- emmeans(m_dmcct, pairwise ~ treatment, adjust = "tukey")

emm_dmcct$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Duration Minimum CCT (Units: $min$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 3e. Diagnostic Plots

```{r}
#residual v fitted plot
resid_dmcct <- tibble(
  fitted = fitted(m_dmcct),
  resid  = resid(m_dmcct)
)

ggplot(resid_dmcct, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Duration Minimum CCT (min)",
    subtitle = expression(paste("Response Units: mins"))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_mcct)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Duration Minimum CCT (min)",
    subtitle = expression(paste("Response Units: mins")),
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
```

# 4. Response = maximum_corneal_pressure

## 4a. Summary Table

```{r}
######
# Response = Total Anesthesia Time
######

#summary table

sum_mp <- df_summ %>%
  group_by(Treatment = treatment) %>%
  summarize(
    N = n(),
    Mean = mean(maximum_corneal_pressure, na.rm = TRUE),
    SD = sd(maximum_corneal_pressure, na.rm = TRUE),
    Min = min(maximum_corneal_pressure, na.rm = TRUE),
    Median = median(maximum_corneal_pressure, na.rm = TRUE),
    Max = max(maximum_corneal_pressure, na.rm = TRUE),
    .groups = "drop"
  )

sum_mp %>%
  kable(
    digits = 2,
    format = "latex",
    caption = paste("Summary of Maximum Corneal Pressure ($g/mm^2$) by Treatment"),
    longtable = FALSE,
    escape = FALSE
  ) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "scale_down"))
```

## 4b. Fit LMM

```{r, results = 'hide'}
#fitting model

m_mp <- lmer(
  maximum_corneal_pressure ~ treatment + period + (1 | horse),
  data = df_summ
)
summary(m_mp)
```

## 4c. Global F Test

```{r}
# capture the ANOVA table as a data.frame
anova_tbl_mp <- as.data.frame(anova(m_mp))

anova_tbl_mp %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>% 
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for Max Corneal Pressure",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 4d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_mp <- emmeans(m_mp, pairwise ~ treatment, adjust = "tukey")

emm_mp$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Maximum Corneal Pressure (Units: $g/mm^2$))",
    longtable = FALSE,
    escape = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 4e. Diagnostic Plots

```{r}
#residual v fitted plot
resid_mp <- tibble(
  fitted = fitted(m_mp),
  resid  = resid(m_mp)
)

ggplot(resid_mp, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Maximum Corneal Pressure",
    subtitle = expression(paste("Response Units: ", g/mm^2))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_tta)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Maximum Conreal Pressure",
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles",
    subtitle = expression(paste("Response Units: ", g/mm^2))
  ) +
  theme_minimal(base_size = 12)
```

# 5. Tweaking the Models

## 5a. Fitting A Log Transform vs Sqrt for total_anesthesia_time_min

```{r}
# 1. Sqrt‐transformed model
m_tta_log1p <- lmer(
  log1p(total_anesthesia_time_min) ~ treatment + (1 | horse),
  data = df_summ, REML = FALSE
)

# 2. Sqrt‐transformed model
m_tta_sqrt <- lmer(
  sqrt(total_anesthesia_time_min) ~ treatment + (1 | horse),
  data = df_summ, REML = FALSE
)

# 3. Diagnostic data frames
diag_log1p <- tibble(
  fitted = fitted(m_tta_log1p),
  resid  = residuals(m_tta_log1p)
)
diag_sqrt <- tibble(
  fitted = fitted(m_tta_sqrt),
  resid  = residuals(m_tta_sqrt)
)

# 4. Plot diagnostics side by side
p1 <- ggplot(diag_log1p, aes(fitted, resid)) +
  geom_point(alpha = .6) +
  geom_hline(yintercept=0, linetype="dashed") +
  labs(title = "Residuals vs Fitted: log1p(TTA)")

p2 <- ggplot(data.frame(resid = residuals(m_tta_log1p)), aes(sample = resid)) +
  stat_qq() + stat_qq_line() +
  labs(title = "QQ-Plot: log1p(TTA)")

p3 <- ggplot(diag_sqrt, aes(fitted, resid)) +
  geom_point(alpha = .6) +
  geom_hline(yintercept=0, linetype="dashed") +
  labs(title = "Residuals vs Fitted: sqrt(TTA)")

p4 <- ggplot(data.frame(resid = residuals(m_tta_sqrt)), aes(sample = resid)) +
  stat_qq() + stat_qq_line() +
  labs(title = "QQ-Plot: sqrt(TTA)")

# Print a 2×2 grid
library(patchwork)
(p1 | p3) / (p2 | p4)
```

## 5b. Global F Test.

```{r}
#########
# WE CHOSE TO WORK WITH THE SQAURE ROOT TRANSFORMATION
########
# 1. Capture the ANOVA table as a data.frame
anova_tbl_sqrt <- as.data.frame(anova(m_tta_sqrt))

anova_tbl_sqrt %>%
  rownames_to_column("Term") %>%
  filter(Term == "treatment") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>%  
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for sqrt(Total Anesthesia Time)",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 5d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_sqrt_tta <- emmeans(m_tta_sqrt, pairwise ~ treatment, adjust = "tukey")

emm_sqrt_tta$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Sqrt Transformed Total Anesthesia Time (Units: $min$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 5e. Diagnostic Plots

```{r}
#residual v fitted plot
resid_sqrt_tta <- tibble(
  fitted = fitted(m_tta_sqrt),
  resid  = resid(m_tta_sqrt)
)

ggplot(resid_sqrt_tta, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Sqrt Transformed Total Anesthesia Time",
    subtitle = expression(paste("Response Units: min"))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_tta_sqrt)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Sqrt Transformed Total Anesthesia Time",
    subtitle = expression(paste("Response Units: min")),
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
```

## 5f. emmip
```{r}
emmip(emm_sqrt_tta, ~ treatment, CI = TRUE)
```

# 6. Adding Horse Period (THIS ONE SHOULD BE INCLUDED IN THE FINAL REPORT)

## 6a. TTA with Horse Period
```{r}

m_tta_hp <- lmer(
  sqrt(total_anesthesia_time_min) ~ treatment +
    (1 | horse) +
    (1 | horse_period),
  data   = df_summ4,
  REML   = TRUE
)

```

## 6b. Global F Test.

```{r}
#########
# WE CHOSE TO WORK WITH THE SQRT TRANSFORMATION
########
# 1. Capture the ANOVA table as a data.frame
anova_tbl_sqrt_hp <- as.data.frame(anova(m_tta_hp))
anova_tbl_sqrt_hp %>%
  rownames_to_column("Term") %>%
  select(
    Term,
    `Num DF`   = NumDF,
    `Den DF`   = DenDF,
    `F value`  = `F value`,
    `Pr(>F)`   = `Pr(>F)`
  ) %>%  
  mutate(
    `Pr(>F)` = if_else(
      `Pr(>F)` < 1e-4,
      formatC(`Pr(>F)`, format = "e", digits = 2),   # scientific for tiny p‐values
      sprintf("%.4f", `Pr(>F)`)                     # 4 decimal places otherwise
    )
  ) %>%
  kable(
    caption = "Type III F‐test for sqrt(Total Anesthesia Time)",
    digits  = c(NA, 0, 0, 3, 6)   # ← last entry = 6 decimals for Pr(>F)
  ) %>%
  kable_styling(full_width = FALSE)
```

## 6d. Pairwise (Tukey) Comnparisons for Treatment

```{r}
#running emmeans
emm_sqrt_tta_hp <- emmeans(m_tta_hp, pairwise ~ treatment, adjust = "tukey")

emm_sqrt_tta_hp$contrasts %>%
  as.data.frame() %>%
  mutate(
    p_value_fmt = if_else(
      p.value < 1e-4,
      formatC(p.value, format = "e", digits = 2),
      sprintf("%.5f", p.value)
    )
  ) %>%
  select(Contrast = contrast, 
         Estimate = estimate,
         SE, 
         DF = df, 
         "T-Ratio" = t.ratio, 
         "P-Value" = p_value_fmt) %>%
  kable(
    digits = c(NA, 3, 2, 0, 2, 4),
    format = "latex",
    caption = "Tukey Pairwise Contrasts for Treatment (Sqrt Transformed Total Anesthesia Time (Units: $min$))",
    longtable = FALSE
  ) %>%
  kable_styling(full_width = TRUE,
                latex_options = c("hold_position", "scale_down"))
```

## 6e. Diagnostic Plots

```{r}
#residual v fitted plot
resid_sqrt_tta_hp <- tibble(
  fitted = fitted(m_tta_hp),
  resid  = resid(m_tta_hp)
)

ggplot(resid_sqrt_tta_hp, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted for Response = Sqrt Transformed Total Anesthesia Time",
    subtitle = expression(paste("Response Units: min"))
  ) +
  theme_minimal(base_size = 12)

#QQ plot
ggplot(data = tibble(resid = resid(m_tta_sqrt)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(
    title = "QQ-plot of Residuals for Response = Sqrt Transformed Total Anesthesia Time",
    subtitle = expression(paste("Response Units: min")),
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)
```

## 6f. emmip
```{r}
emmip(emm_sqrt_tta_hp, ~ treatment, CI = TRUE)
```

## 6g. Is Adding horse_period significant?
```{r}
m_base_ml <- lmer(
  log1p(total_anesthesia_time_min) ~ treatment + (1 | horse),
  data = df_summ4,
  REML = FALSE
)

# 1. Get the comparison table
lrt_tbl <- as.data.frame(anova(m_base_ml, m_tta_hp))

# 2. Bring the rownames in as a column named "Model"
lrt_tbl <- lrt_tbl %>%
  rownames_to_column(var = "Model")

# 3. Select & rename the relevant columns
lrt_tbl %>%
  select(
    Model,
    AIC,
    BIC,
    LogLik    = logLik,
    Chisq,
    Df,
    `p-value` = `Pr(>Chisq)`
  ) %>%
  kable(
    caption = "Likelihood‐Ratio Test for Adding horse\\_period Random Effect",
    digits  = c(NA, 1, 1, 2, 2, 0, 3)
  ) %>%
  kable_styling(full_width = FALSE)
```



## 7a. Fitting the New Model, with Both HemScore and Horse Period
```{r}
m_tta_hp_hs <- lmer(
  log1p(total_anesthesia_time_min) ~ treatment + hem_score +
    (1 | horse) + (1 | horse_period),
  data = df_summ4,
  REML = TRUE
)
```

```{r}
#testing whether hemscore helps
print(VarCorr(m_tta_hp_hs), comp = c("Variance","Std.Dev"))

anova(m_tta_hp_hs, type="III")

```
